### 2019-06-13 2-Chapters/3-Chapter/chapter3.tex
Last proof reading of Ch6 and conclusion

SMPyBandits stands for \emph{S}ingle and \emph{M}ulti \emph{P}layer \emph{Bandits} in P\emph{y}thon.
The library does not aim at being blazing fast or optimized in terms of memory usage, as it comes with a pure Python implementation \cite{python}, and uses only standard open-source Python packages (\ie, no hard to install dependencies).
Some critical parts are also available as a \texttt{C} Python extension, and the just-in-time Numba compiler \cite{numba} is used whenever it is possible, so we can note that we optimized the time efficiency of what could be (easily) optimized.
However if simulation speed really matters, one should rather refer to less exhaustive but faster implementations, like for example \cite{TorLibbandit} in \texttt{C++} or \cite{VishMABjl} in Julia. Note that both are not maintained anymore, and contain just a few algorithms for the simple stationary MAB model.

@@

\subsection{Reward distributions}
%
We focus on one dimensional distributions, implemented in the \texttt{\href{https://smpybandits.github.io/docs/Arms.html}{Arms}} module.
The library supports discrete distributions: \emph{Bernoulli} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Bernoulli.html}{Bernoulli}}), \emph{binomial} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Binomial.html}{Binomial}}), \emph{Poisson} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Poisson.html}{Poisson}}), and a generic \emph{discrete} distribution (\texttt{\href{https://smpybandits.github.io/docs/Arms.DiscreteArm.html}{DiscreteArm}}),
as well as continuous distributions,
which can be truncated to an interval $[a,b]$ or have unbounded support ($\mathbb{R}$):
\emph{exponential} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Exponential.html}{Exponential}}), \emph{gamma} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Gamma.html}{Gamma}}), \emph{Gaussian} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Gaussian.html}{Gaussian}}) and \emph{uniform} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Uniform.html}{Uniform}}).
%
The default is to use the same distribution for the $K$ arms, as this is the setting studied in the literature,
but it also possible to mix them.
%  (\eg, one arm is Bernoulli-distributed and another).
%  (see for instance Figure~\ref{fig:25:HarderMixed}).
%
For instance the following code in Code~\ref{lst:3:pythonCodeCreateProblem} creates three arms, following a Bernoulli distribution with respective means $0.1, 0.5, 0.9$, and a \texttt{MAB} object which encapsulate sthe problem.
We give another example for truncated Gaussian distributions, and a visualization of a histogram of $10000$ rewards obtained from the arms, below in Figure~\ref{fig:3:exampleOfRewards}.

@@

Simulations are easily executed, \eg, Code~\ref{lst:3:howToRunBasicLibrary} shows how to start $N=1000$ repetitions of a simple non-Bayesian Bernoulli-distributed problem, for $K=9$ arms, an horizon of $T=10000$ and on $4$ CPUs.
It takes about $20$ min, on a $4$-core $64$-bit GNU/Linux laptop.
Environment variables (\texttt{N=1000} etc) in the command line are not required, but they are convenient.

@@

    \item
    %  Algorithm 8 is
    AdBandits from \cite{Truzzi13} (\texttt{AdBandits}), using $\alpha=1$.
    % It follows the same assumptions as Thompson sampling and Bayes-UCB, and uses a Beta posterior on each arm. It uses a random mixture between the two algorithms, and with a certain probability it uses the Thompson decision rule (\ie, choose the arm maximizing one draw of random samples from the posteriors), otherwise it uses the Bayes-UCB decision rule (\ie, choose the arm maximizing the $1-1/t$ quantile).
    % It was proposed in this article \cite{Truzzi13} and illustrated on empirical simulations, but no theoretical analysis was given, and it did not gain popularity since then.
    In practice, it is usually (slighly) more efficient than both Thompson sampling and Bayes-UCB in terms of regret, it is comparable in terms of computation times, but costs more memory.

>>>>

SMPyBandits stands for \emph{S}ingle- and \emph{M}ulti-\emph{P}layer \emph{Bandits} in P\emph{y}thon.
The library does not aim at being blazing fast or optimized in terms of memory usage, as it comes with a pure Python implementation \cite{python}, and uses only standard open-source Python packages (\ie, no hard to install dependencies).
Some critical parts are also available as a \texttt{C} Python extension, and the just-in-time Numba compiler \cite{numba} is used whenever it is possible, so we can note that we optimized the time efficiency of what could be (easily) optimized.
However if simulation speed really matters, one should rather refer to less exhaustive but faster implementations, like for example \cite{TorLibbandit} in \texttt{C++} or \cite{VishMABjl} in Julia. Note that both are not maintained anymore, and contain just a few algorithms for the simple stationary MAB model.

@@

\subsection{Reward distributions}
%
We focus on one dimensional distributions, implemented in the \texttt{\href{https://smpybandits.github.io/docs/Arms.html}{Arms}} module.
The library supports discrete distributions: \emph{Bernoulli} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Bernoulli.html}{Bernoulli}}), \emph{binomial} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Binomial.html}{Binomial}}), \emph{Poisson} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Poisson.html}{Poisson}}), and a generic \emph{discrete} distribution (\texttt{\href{https://smpybandits.github.io/docs/Arms.DiscreteArm.html}{DiscreteArm}}),
as well as continuous distributions,
which can be truncated to an interval $[a,b]$ or have unbounded support ($\mathbb{R}$):
\emph{exponential} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Exponential.html}{Exponential}}), \emph{gamma} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Gamma.html}{Gamma}}), \emph{Gaussian} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Gaussian.html}{Gaussian}}) and \emph{uniform} (\texttt{\href{https://smpybandits.github.io/docs/Arms.Uniform.html}{Uniform}}).
%
The default is to use the same distribution for the $K$ arms, as this is the setting studied in the literature,
but it also possible to mix them.
%  (\eg, one arm is Bernoulli-distributed and another).
%  (see for instance Figure~\ref{fig:25:HarderMixed}).
%
For instance the following code in Code~\ref{lst:3:pythonCodeCreateProblem} creates three arms, following a Bernoulli distribution with respective means $0.1, 0.5, 0.9$, and a \texttt{MAB} object which encapsulates the problem.
We give another example for truncated Gaussian distributions, and a visualization of a histogram of $10000$ rewards obtained from the arms, below in Figure~\ref{fig:3:exampleOfRewards}.

@@

Simulations are easily executed, \eg, Code~\ref{lst:3:howToRunBasicLibrary} shows how to start $N=1000$ repetitions of a simple non-Bayesian Bernoulli-distributed problem, for $K=9$ arms, an horizon of $T=10000$ and on $4$ CPU.
It takes about $20$ min, on a $4$-core $64$-bit GNU/Linux laptop.
Environment variables (\texttt{N=1000} etc) in the command line are not required, but they are convenient.

@@

    \item
    %  Algorithm 8 is
    AdBandits from \cite{Truzzi13} (\texttt{AdBandits}), using $\alpha=1$.
    % It follows the same assumptions as Thompson sampling and Bayes-UCB, and uses a Beta posterior on each arm. It uses a random mixture between the two algorithms, and with a certain probability it uses the Thompson decision rule (\ie, choose the arm maximizing one draw of random samples from the posteriors), otherwise it uses the Bayes-UCB decision rule (\ie, choose the arm maximizing the $1-1/t$ quantile).
    % It was proposed in this article \cite{Truzzi13} and illustrated on empirical simulations, but no theoretical analysis was given, and it did not gain popularity since then.
    In practice, it is usually (slightly) more efficient than both Thompson sampling and Bayes-UCB in terms of regret, it is comparable in terms of computation times, but costs more memory.

### 2019-06-13 2-Chapters/5-Chapter/ALT_2018__MPBandits.git/mpBandits.tex
Last proof reading of Ch6 and conclusion

\begin{leftbar}[warningbar]  % XXX leftbar warningbar, comment if needed
  % FIXME
  \textbf{\textcolor{red}{Warning}.}
  %
  The regret lower bound that we gave in \cite{Besson2018ALT} is not applicable to any algorithm,
  as it was discovered by Etienne Boursier and Vianney Perchet, as they explain in Section~2.4 in \cite{BoursierPerchet18}.
  The proof we gave in the Appendix of our paper \cite{Besson2018ALT} was wrong in just one step,
  but the results stated in this section are now correct,
  thanks to the modifications proposed by Emilie Kaufmann and Abbas Mehrabian in Appendix~E of \cite{KaufmannAbbas19}.
  The cost for this modification is to restrict the lower-bound to a smaller class of algorithms, which \emph{should} contain \RhoRand, \RandTopM{} and \MCTopM, but not the two algorithms proposed SIC-MMAB from \cite{BoursierPerchet18} and Multiplayer Explore-then-Commit (M-ETC) from \cite{KaufmannAbbas19}.
  Proving that the proposed algorithms are in this class is left as an open question.
\end{leftbar}  % XXX leftbar warningbar, comment if needed

@@

\begin{lemma}\label{lem:5:1stLowerBound}
\begin{leftbar}[lemmabar]  % XXX leftbar lemmabar, comment if needed
    For any strategy $\cA$ and $\boldsymbol{\mu}\in\cP_M$, the regret is lower-bounded by \ref{eq:5:term1}:
    \begin{align*}
        R_T(\boldsymbol{\mu}, M, \cA)    \geq \sum\limits_{k \in \Mworst} (\mu_M^*- \mu_k) \E_{\mu}[N_k(T)].
    \end{align*}
\end{leftbar}  % XXX leftbar lemmabar, comment if needed
\end{lemma}

\begin{smallproof}
  Note that term \ref{eq:5:term3} is clearly lower bounded by $0$
  but it is not obvious for \ref{eq:5:term2} as there is no reason for $N_k(T)$ to be upper bounded by $T$.
  %
  Let $N_k^{!}(T) \eqdef \sum_{t=1}^{T} \indic(\exists! j, A^j(t)=k)$,
  where the notation $\exists!$ stands for ``there exists a unique''.
  Then $N_k(T) = \sum_{t=1}^{T} \sum_{j=1}^{M} \indic(A^j(t) = k)$ can be decomposed as
  \begin{equation*}
    N_k(T) = \sum_{t=1}^{T} \indic(\exists! j, A^j(t) = k) + \sum_{t=1}^{T} \sum_{j=1}^{M} c_{k,t} \indic(A^j(t) = k)
    = N_k^{!}(T) + C_k(T).
  \end{equation*}
  By focusing on the two terms $\ref{eq:5:term2} + \ref{eq:5:term3}$ from the decomposition of $R_T(\boldsymbol{\mu}, M, \cA)$ from Lemma~\ref{lem:5:DecompositionRegret}, we have
  \begin{align*}
    \ref{eq:5:term2} + \ref{eq:5:term3} &=
    \sum_{k \in \Mbest} (\mu_k - \mu_M^*) (T - \E_{\mu}[N_k^!(T)])
    + \sum_{k \in \Mbest} \mu_M^* \E_{\mu}[C_k(T)] \\
    & \;\;\;\;\;\; + \sum_{k=1}^{M} \mu_k \E_{\mu}[C_k(T)]
    - \sum_{k \in \Mbest} \mu_k \E_{\mu}[C_k(T)] \\
    &=
    \sum_{k \in \Mbest} (\mu_k - \mu_M^*) (T - \E_{\mu}[N_k^!(T)])
    + \sum_{k \in \Mbest} \mu_M^* \E_{\mu}[C_k(T)]
    + \sum_{k \in \Mworst} \mu_k \E_{\mu}[C_k(T)] \\
    &=
    \sum_{k \in \Mbest} (\mu_k - \mu_M^*) (T - \E_{\mu}[N_k^!(T)])
    + \sum_{k=1}^{M} \min(\mu_M^*, \mu_k) \E_{\mu}[C_k(T)].
  \end{align*}
  And now both terms are non-negative, as $N_k^!(T) \leq T$, $\min(\mu_M^*, \mu_k)\geq 0$, and $C_k(T) \geq 0$, so $\ref{eq:5:term2} + \ref{eq:5:term3} \geq 0$
  which proves that $R_T(\boldsymbol{\mu}, M, \cA) \geq \ref{eq:5:term1}$, as wanted.
\end{smallproof}

@@

\paragraph{Logarithmic switching cost for \MCTopM.}\label{app:5:NumberSwitches}
%
While Lemma~\ref{lem:5:collisionsMCTopM} bounds the number of collisions,
another consequence of our proof is that we have also bounded
the (expected) number of \emph{switches of arms}, $\mathrm{SC}^{\cA}(T) = \sum_{t=1}^{T-1} \Pr(A^j(t+1) \neq A^j(t))$.
%
We controlled the total number of transitions $(2)$, $(3)$ and $(5)$ (see Figure~\ref{fig:5:StateMachineAlgorithm_MCTopM}),
which are the only transitions when a player can switch from arm $k$ to arm $k'\neq k$.
Thus, the total number of arm switches is also proven to be logarithmic, if all players uses the \MCTopM-\klUCB{} algorithm.
This additional guarantee was never clearly stated for previous state-of-the-art works, like \rhoRand.
%
Even though minimizing the number of arms switching was not a goal,
this guarantee is appealing, in particular for Cognitive Radio applications,
where switching arms means re-configuring a radio hardware, an operation that costs energy.
An algorithm guaranteeing a small number of switches is thus interesting,
and for instance \cite{modiDemo2016} studied the empirical impact of the number of hardware reconfigurations on the battery life of a dynamic end-devices.
This work highlighted a tradeoff between minimizing regret and minimizing the number of switches of arms.
%
We also note that introducing explicit \emph{switching costs} in the regret, like it was done in previous works like \cite{Koren17} for single-player bandits, could also be fruitful.

>>>>

\begin{leftbar}[warningbar]  % XXX leftbar warningbar, comment if needed
  % FIXME
  \textbf{\textcolor{red}{Warning}.}
  %
  The regret lower bound that we gave in \cite{Besson2018ALT} is not applicable to any algorithm,
  as it was discovered by Boursier and Perchet, as they explain in Section~2.4 in \cite{BoursierPerchet18}.
  The proof we gave in the Appendix of our paper \cite{Besson2018ALT} was wrong in just one step,
  but the results stated in this section are now correct,
  thanks to the modifications proposed by Kaufmann and Mehrabian in Appendix~E of \cite{KaufmannAbbas19}.
  The cost for this modification is to restrict the lower-bound to a smaller class of algorithms, which \emph{should} contain \RhoRand, \RandTopM{} and \MCTopM, but not the two algorithms proposed SIC-MMAB from \cite{BoursierPerchet18} and Multiplayer Explore-then-Commit (M-ETC) from \cite{KaufmannAbbas19}.
  Proving that the proposed algorithms are in this class is left as an open question.
\end{leftbar}  % XXX leftbar warningbar, comment if needed

@@

\begin{lemma}\label{lem:5:1stLowerBound}
\begin{leftbar}[lemmabar]  % XXX leftbar lemmabar, comment if needed
    For any strategy $\cA$ and $\boldsymbol{\mu}\in\cP_M$, the regret is lower-bounded by \ref{eq:5:term1}:
    \begin{align*}
        R_T(\boldsymbol{\mu}, M, \cA)    \geq \sum\limits_{k \in \Mworst} (\mu_M^*- \mu_k) \E_{\mu}[N_k(T)].
    \end{align*}%
\end{leftbar}  % XXX leftbar lemmabar, comment if needed
\end{lemma}

\begin{smallproof}
  Note that term \ref{eq:5:term3} is clearly lower bounded by $0$
  but it is not obvious for \ref{eq:5:term2} as there is no reason for $N_k(T)$ to be upper bounded by $T$ (it counts the selections of arm $k$ by \emph{all} the players, and for instance if player $1$ is fixed on arm $1$ and player $2$ plays it at least once, then $N_1(T)\geqT+1$).
  %
  Let $N_k^{!}(T) \eqdef \sum_{t=1}^{T} \indic(\exists! j, A^j(t)=k)$,
  where the notation $\exists!$ stands for ``there exists a unique''.
  Then $N_k(T) = \sum_{t=1}^{T} \sum_{j=1}^{M} \indic(A^j(t) = k)$ can be decomposed as
  \begin{equation*}
    N_k(T) = \sum_{t=1}^{T} \indic(\exists! j, A^j(t) = k) + \sum_{t=1}^{T} \sum_{j=1}^{M} c_{k,t} \indic(A^j(t) = k)
    = N_k^{!}(T) + C_k(T).
  \end{equation*}
  By focusing on the two terms $\ref{eq:5:term2} + \ref{eq:5:term3}$ from the decomposition of $R_T(\boldsymbol{\mu}, M, \cA)$ from Lemma~\ref{lem:5:DecompositionRegret}, we have
  \begin{align*}
    \ref{eq:5:term2} + \ref{eq:5:term3} &=
    \sum_{k \in \Mbest} (\mu_k - \mu_M^*) (T - \E_{\mu}[N_k^!(T)])
    + \sum_{k \in \Mbest} \mu_M^* \E_{\mu}[C_k(T)] \\
    & \;\;\;\;\;\; + \sum_{k=1}^{M} \mu_k \E_{\mu}[C_k(T)]
    - \sum_{k \in \Mbest} \mu_k \E_{\mu}[C_k(T)] \\
    &=
    \sum_{k \in \Mbest} (\mu_k - \mu_M^*) (T - \E_{\mu}[N_k^!(T)])
    + \sum_{k \in \Mbest} \mu_M^* \E_{\mu}[C_k(T)]
    + \sum_{k \in \Mworst} \mu_k \E_{\mu}[C_k(T)] \\
    &=
    \sum_{k \in \Mbest} (\mu_k - \mu_M^*) (T - \E_{\mu}[N_k^!(T)])
    + \sum_{k=1}^{M} \min(\mu_M^*, \mu_k) \E_{\mu}[C_k(T)].
  \end{align*}
  And now both terms are non-negative, as $N_k^!(T) \leq T$, $\min(\mu_M^*, \mu_k)\geq 0$, and $C_k(T) \geq 0$, so $\ref{eq:5:term2} + \ref{eq:5:term3} \geq 0$
  which proves that $R_T(\boldsymbol{\mu}, M, \cA) \geq \ref{eq:5:term1}$, as wanted.
\end{smallproof}

@@

\paragraph{Logarithmic switching cost for \MCTopM.}\label{app:5:NumberSwitches}
%
While Lemma~\ref{lem:5:collisionsMCTopM} bounds the number of collisions,
another consequence of our proof is that we have also bounded
the (expected) number of \emph{switches of arms}, $\mathrm{SC}^{\cA}(T) = \sum_{t=1}^{T-1} \Pr(A^j(t+1) \neq A^j(t))$.
%
We controlled the total number of transitions $(2)$, $(3)$ and $(5)$ (see Figure~\ref{fig:5:StateMachineAlgorithm_MCTopM}),
which are the only transitions when a player can switch from arm $k$ to arm $k'\neq k$.
Thus, the total number of arm switches is also proven to be logarithmic, if all players uses the \MCTopM-\klUCB{} algorithm.
This additional guarantee was never clearly stated for previous state-of-the-art works, like \rhoRand.
%
Even though minimizing the number of arms switching was not a goal,
this guarantee is appealing, in particular for Cognitive Radio applications,
where switching arms means re-configuring a radio hardware, an operation that costs energy.
An algorithm guaranteeing a small number of switches is thus interesting,
and for instance \cite{modiDemo2016} studied the empirical impact of the number of hardware reconfigurations on the battery life of a dynamic end-devices.
This work highlighted a tradeoff between minimizing regret and minimizing the number of switches of arms.
On a more experimental note, it was shown in \cite{modiDemo2016} that the previous state-of-the-art policy for multi-players bandits, \rhoRand, could be tuned to run in batches, in order to reduce by a certain multiplicative factor its switching cost while only adding an additive factor on its regret.
While such ideas can interest a practitioner, they does not change the asymptotic behavior of $\mathrm{SC}^{\cA_1,\dots,\cA_M}(T)$, which is $\Theta(\log(T))$ for any efficient policy.
%
We also note that introducing explicit \emph{switching costs} in the regret, like it was done in previous works like \cite{Koren17} for single-player bandits, could also be fruitful.

% On a more experimental note, it was shown in \cite{modiDemo2016} that the previous state-of-the-art policy for multi-players bandits, \rhoRand, could be tuned to run in batches\footnote{~The batch bandit setting implies that all players use the same decisions for instance for $50$ consecutive times, and update their decisions only once every $50$ time steps. For more details, see \cite{modiDemo2016} for experiments on the multi-players case, or \cite{perchet2016,gao2019batched,kolnogorov2019multi} for theoretical developments on the single-player case.}, in order to reduce by a certain multiplicative factor its switching cost while only adding an additive factor on its regret. While such ideas can interest a practitioner, they does not change the asymptotic behavior of $\mathrm{SC}^{\cA_1,\dots,\cA_M}(T)$, which is $\Theta(\log(T))$ for any efficient policy.
